---
title: "Homework2"
output: github_document
---

## Library

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r}
NY_transit_df =
  read.csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("",".","NA")
  ) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type,ada
  ) |>
  mutate(entry = case_match(
    entry,
    "YES" ~ TRUE,
    "NO" ~ FALSE))
```

### Write a short paragraph about this dataset

This dataset contains information about subway entrances and exits in NYC. The variables include line, station name, station latitude and longitude, routes served, entry, vending, entrance type, and ADA compliance.
I cleaned the column names, kept the required variables, and changed `entry` from YES/NO to TRUE/FALSE.
The cleaned dataset has `r nrow(NY_transit_df)` rows and `r ncol(NY_transit_df)` columns. These data are not fully tidy because route names are stored in separate columns, from `route1` to `route11`.

### Answer the following questions using these data

```{r}
station_count =
  NY_transit_df |>
  distinct(line, station_name) |>
  nrow()

station_count
```
There are r station_count distinct stations, identified by line and station name.

```{r}
ada_station_count =
  NY_transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

ada_station_count
```
There are r ada_station_count ADA-compliant stations.

```{r}
no_vending_summary =
  NY_transit_df |>
  filter(vending == "NO") |>
  summarise(
    total = n(),
    allow_entry = sum(entry),
    proportion = mean(entry)
  )

no_vending_summary
```
Among entrances and exits without vending, r round(no_vending_summary$proportion * 100, 2)% allow entry.

### Reformat data and answer

```{r}
NY_transit_tidy_df =
  NY_transit_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )
```

```{r}
a_station_count =
  NY_transit_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_station_count
```
After reshaping the route columns into route_number and route_name, there are r a_station_count distinct stations that serve the A train.

```{r}
a_ada_station_count =
  NY_transit_tidy_df |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()

a_ada_station_count
```
Of the stations that serve the A train, r a_ada_station_count are ADA compliant.


## Problem 2

### Clean Mr. Trash Wheel

```{r}
mr_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel", range = "A2:N742"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Mr. Trash Wheel"
  )
```

### Clean Professor Trash Wheel

```{r}
professor_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    plastic_bottles = as.numeric(
      na_if(plastic_bottles, "dive")),
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel"
  )
```

### Clean Gwynnda

```{r}
gwynnda_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Gwynnda"
  )
```

### Combine these datasets

```{r}
trash_wheel_df =
  bind_rows(
    mr_trash_df,
    professor_trash_df,
    gwynnda_trash_df
  )
```


```{r}
professor_total_weight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum()
professor_total_weight

gwynnda_june_cigarettes =
  trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  pull(cigarette_butts) |>
  sum()
gwynnda_june_cigarettes
```

The combined dataset contains `r nrow(trash_wheel_df)` observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Each observation is a dumpster record. 
Key variables include the date, weight in tons, volume in cubic yards, and counts of items such as plastic bottles and cigarette butts. The `trash_wheel` variable shows which wheel each record comes from. Some variables are not available for all three wheels and are recorded as `NA`. 
For the available data, Professor Trash Wheel collected a total of `r round(professor_total_weight, 2)` tons of trash. Gwynnda collected `r gwynnda_june_cigarettes` cigarette butts in June 2022.

## Problem 3

```{r}
baseline_raw_df =
  read_csv(
    "data_mci/MCI_baseline.csv",
    skip = 1,
    na = c("", ".", "NA")
  ) |>
  janitor::clean_names()

baseline_df =
  baseline_raw_df |>
  rename(
    baseline_age = current_age,
    mci_onset_age = age_at_onset
  ) |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier"))
  ) |>
  filter(
    is.na(mci_onset_age) | mci_onset_age > baseline_age)

glimpse(baseline_df)
```

```{r}
recruited_count = nrow(baseline_raw_df)

eligible_count = nrow(baseline_df)

excluded_count = recruited_count - eligible_count

mci_count =
  baseline_df |>
  summarise(
    total = sum(!is.na(mci_onset_age))
  ) |>
  pull(total)

average_baseline_age =
  baseline_df |>
  pull(baseline_age) |>
  mean()

women_carrier_proportion =
  baseline_df |>
  filter(sex == "Female") |>
  summarise(
    proportion = mean(apoe4 == "Carrier")
  ) |>
  pull(proportion)
```

The baseline dataset contains participants' ID, baseline age, sex, education, APOE4 carrier status, and age at MCI onset. I skipped the first row because it contains variable descriptions and treated "." as a missing value. I renamed the age variables and converted sex and APOE4 status to factors. I removed participants whose MCI onset age was less than or equal to their baseline age. Participants with missing onset age were retained because they remained MCI-free during follow-up.

The original dataset contains `r recruited_count` participants. After excluding `r excluded_count` participants, `r eligible_count` remain, with one row per participant. Of these, `r mci_count` developed MCI during follow-up. The average baseline age is `r round(average_baseline_age, 2)` years. Among women, `r round(women_carrier_proportion * 100, 1)`% are APOE4 carriers.





