---
title: "Homework2"
output: github_document
---

## Library

```{r}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r}
NY_transit_df =
  read.csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("",".","NA")
  ) |>
  janitor::clean_names() |>
  select(line, station_name, station_latitude, station_longitude,
         route1:route11, entry, vending, entrance_type,ada
  ) |>
  mutate(entry = case_match(
    entry,
    "YES" ~ TRUE,
    "NO" ~ FALSE))
```

### Write a short paragraph about this dataset

This dataset contains information about subway entrances and exits in NYC. The variables include line, station name, station latitude and longitude, routes served, entry, vending, entrance type, and ADA compliance.
I cleaned the column names, kept the required variables, and changed `entry` from YES/NO to TRUE/FALSE.
The cleaned dataset has `r nrow(NY_transit_df)` rows and `r ncol(NY_transit_df)` columns. These data are not fully tidy because route names are stored in separate columns, from `route1` to `route11`.

### Answer the following questions using these data

```{r}
station_count =
  NY_transit_df |>
  distinct(line, station_name) |>
  nrow()

station_count
```
There are r station_count distinct stations, identified by line and station name.

```{r}
ada_station_count =
  NY_transit_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

ada_station_count
```
There are r ada_station_count ADA-compliant stations.

```{r}
no_vending_summary =
  NY_transit_df |>
  filter(vending == "NO") |>
  summarise(
    total = n(),
    allow_entry = sum(entry),
    proportion = mean(entry)
  )

no_vending_summary
```
Among entrances and exits without vending, r round(no_vending_summary$proportion * 100, 2)% allow entry.

### Reformat data and answer

```{r}
NY_transit_tidy_df =
  NY_transit_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    cols = route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name",
    values_drop_na = TRUE
  ) |>
  mutate(
    route_number = as.integer(route_number)
  )
```

```{r}
a_station_count =
  NY_transit_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_station_count
```
After reshaping the route columns into route_number and route_name, there are r a_station_count distinct stations that serve the A train.

```{r}
a_ada_station_count =
  NY_transit_tidy_df |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()

a_ada_station_count
```
Of the stations that serve the A train, r a_ada_station_count are ADA compliant.


## Problem 2

### Clean Mr. Trash Wheel

```{r}
mr_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel", range = "A2:N742"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Mr. Trash Wheel"
  )
```

### Clean Professor Trash Wheel

```{r}
professor_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    plastic_bottles = as.numeric(
      na_if(plastic_bottles, "dive")),
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel"
  )
```

### Clean Gwynnda

```{r}
gwynnda_trash_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    month = str_trim(month),
    year = as.integer(year),
    trash_wheel = "Gwynnda"
  )
```

### Combine these datasets

```{r}
trash_wheel_df =
  bind_rows(
    mr_trash_df,
    professor_trash_df,
    gwynnda_trash_df
  )
```







